Evaluating the Capabilities of Lattice Boltzmann Method for non-Newtonian and Free-surface Flows towards Applications in Wellbore Cementing

### 2016-07-28 thesis.tex
wrote some basic algorithms
finished first round of revisions

%\usepackage{amsthm}
\usepackage{amssymb}
\usepackage{mathtools}
\usepackage{bm}
\usepackage[plain]{fancyref}
\usepackage{tabulary}
\usepackage{pbox}
\usepackage{algorithm}
\usepackage[noend]{algpseudocode}

@@

\tableofcontents
\listoftables
\listoffigures

@@

When oil and gas wellbores are drilled, barriers must be put in place to ensure that fluids do not leak out of the wellbore.
Wellbore leakage can lead to environmental damage, loss of pressure at the wellhead, and consequently, loss of production.
An important yet vulnerable barrier is the cement annulus.
Creating the cement annulus, a process known as primary cementing, is difficult to perform optimally.
Every well has unique subsurface conditions, and so no cement slurry mix design both performs well and is economical for all wells.
Although some general guidelines and analytical techniques exist for approximating the performance of a cement slurry mix, the mechanics of primary cementing are complex.
Computational methods can help better understand primary cementing and aid designers in determining the optimal mix.
The lattice Boltzmann method (LBM) is a promising technique for simulating primary cementing because it is well-suited for efficiently simulating non-Newtonian flows, multiphase multicomponent flows, and flows through complex geometries--namely, some of the complexities associated with the mechanics of primary cementing.
In \Fref{sec:numerical-methods}, an algorithm for simulating non-Newtonian free-surface flow using LBM is presented.
The algorithm was implemented and used to study primary cementing of a dry annulus, i.e. an annulus that is not filled with drilling mud.
More specifically, the study involved parameterizing different cement slurry flows and investigating how well each slurry flow filled different wellbore geometries.
The study is followed by conclusions and a discussion of future work.

@@

Of critical importance to drilling operations of all kinds is that barriers are placed to prevent gasses and fluids from migrating from one geological zone to another (i.e., zonal isolation is achieved).
If zonal isolation is not achieved, gasses and fluids may migrate through the well to the surface and up into the atmosphere--potentially causing pollution, or at the very least, contributing to an increase in greenhouse gasses--or, worse yet, formation gasses and fluids can migrate into aquifers, potentially harming wild life and people.
Furthermore, it can be argued that the most important, yet vulnerable, barrier in terms of achieving zonal isolation of a wellbore is the cement annulus that is created between the casing and the formation.
There are various mechanisms by which this cement annulus can fail.
According to previous studies~\cite{Bit02,Che85,Lev79,Ste88,Tal93,Zul12}, the most prevalent failure mechanisms of the cement annulus are:
\begin{itemize}
\item Stresses developing in the cement annulus as a result of temperature gradients, moisture gradients, overburden pressure, and chemical shrinkage of the cement matrix.
These stresses can cause cracking in the cement, or debonding at either the casing--cement or cement--formation interfaces.
\item Poor fill of the annulus. If cement slurry does not fill the annulus then if drilling mud is left behind it can dry, crack, and provide a weak path for fluids and gasses under pressure to push through, or if a dry hole is is cemented and not filled properly, fluids and gasses can travel through channels and voids left in the annular space.
\item Gas channeling through the annulus during cement curing.
Initially, after the annulus is poured, the cement column transmits its full hydrostatic pressure against the rock formation.
As the cement annulus hydrates, it solidifies, becoming stiffer and more self-supporting.
When the column begins to support itself, the hydrostatic pressure of the cement column--against the rock formation--begins to drop.
If the hydrostatic pressure of the cement column drops below the pressure of a formation gas before the cement annulus has developed adequate strength to resist it, the formation gas can penetrate int the cement annulus and form channels, or degrade the cement annulus in other ways.
\end{itemize}
% what research has been done that is considered classical?

@@

In order to better understand, and thus improve primary cementing for zonal isolation, several works have attempted to model certain aspects of the multiphase and multiphysics processes of wellbore cement placement.
For example,~\citet{Bei77} developed a mathematical model to ``describe the miscible displacement of drilling muds by cement slurries under laminar flow conditions''.
Among their assumptions were that the process could be modeled in a quasi-steady manner, the volumetric flow rate was constant at each cross section of the flow, and the pressure fields were only dependent on the y-coordinate.
The model was then used to explore the effect of various parameters on the displacement efficiency of the spacer, such as the density ratio between the mud and the spacer, the viscosity ratio between the mud and the spacer, the displacement rate, the displaced phase yield stress, and the displacing phase yield stress.
It was concluded that displacement efficiency could be improved by increasing the density of the displacing phase, increasing the viscosity of the displacing phase, (in general) decreasing the displacement rate, decreasing the yield stress of the displaced phase, and increasing the yield stress of the displacing phase.
\citet{Bit02} developed a mathematical model for primary cementing of an oil well using a Hele-Shaw displacement model.
The model was more general than the model developed by~\citeauthor{Bei77} as it used less restrictive assumptions and was able to consider an eccentric annulus.
Annulus eccentricity occurs often in wellbores and can sometimes be extreme as it is difficult to keep long strings of casing centered in an imperfectly drilled hole.
The bulk fluid motion of the spacer and drilling mud during primary cementing were investigated.
The intended use of the model was not to make general statements about what properties of the spacer and drilling mud were desirable, but to instead be used iteratively during the design process for the cement job.
Results showed that the model met its intended purpose, and in particular, was able to simulate an unyielded channel of mud left on the narrow side of the annulus.
\citet{Zul12} used a computational fluids dynamics (CFD) solver with volume of fluid (VOF) method to simulate the cementing process of a 50 ft. section of wellbore.
The model included the interfacial dynamics of, not just the spacer and drilling mud, but also the cement slurry--which was not considered in either~\citeauthor{Bei77} or~\citeauthor{Bit02}.
The drilling mud and cement slurry properties were kept constant, while the spacer density, viscosity, and displacement rates were varied.
\citeauthor{Zul12} concluded that the high displacement efficiency occurred when the spacer was the same density as the drilling mud and had a smaller viscosity, similar to that of water.
The displacement efficiency was poor for all scenarios in which the spacer and cement slurry densities were equal.
\citet{zhao2016lattice} developed a model simulating primary cementing using the Lattice Boltzmann Method.
\citeauthor{zhao2016lattice} used the model to investigate the displacement interface shape in a horizontal, eccentric annulus; and in order to simplify the model from a 3D to 2D flow, used the classical Hele-Shaw approach.
It was found, in general, that a greater density ratio between the displacing phase, i.e. cement slurry, and displaced phase, i.e. drilling mud, and lower eccentricity, results in better displacement in a horizontal, eccentric annulus.

@@

The purpose of the current work is to develop a computational approach to investigate the flow of cement slurry in and around the--possibly--imperfect features of a rock formation in a vertical section of a wellbore, with the hope of better understanding what parameters of the primary cement job will lead to a good fill of the cement annulus and high bond strength at imperfect rock formation features (i.e., the purpose of the current work is an extension of the work present in~\citet{grasinger2015simulation}).
The Lattice Boltzmann Method (LBM) was chosen as the most suitable modeling technique for the desired purpose because of its ability to model flow through and around complex geometries~\cite{thorne2006lattice}, such as those found at the surface of the rock formation, among other benefits.
LBM also has other advantages as an approach for simulating primary cementing:
\begin{itemize}
\item LBM is well-suited for simulating non-Newtonian fluids because the strain-rate tensor is computed local to each node and is second-order accurate~\cite{kruger2009shear,kruger2010second}.
Many fluids present in primary cementing such as the spacer, drilling mud, and most importantly, the cement slurry, are best modeled as non-Newtonian fluids.
\item LBM is well-suited for simulating multiphase multicomponent flows and free-surface flows because it can handle complicated interface shapes between fluid phases and components.
\item LBM is easily written in parallel.
As hardware architectures shift from one or a few cores to, especially in the case of scientific computing, several CPUs, it becomes increasingly important that code for simulating complex and computationally expense phenomena be written in parallel.
\end{itemize}
The current work focuses on the problem of cementing a dry hole, i.e. an annulus that is not filled with drilling mud prior to cementing.
In order to numerically investigate this problem, an LBM model for non-Newtonian and free-surface flow is developed.
The details of the LBM model are outlined in \Fref{sec:numerical-methods}.
What follows in \Fref{sec:results} is a preliminary study on how different characteristics of flow, e.g. cement slurry yield stress, cement slurry inlet velocity, etc., perform in regards to fill of the annulus for different wellbores with imperfect rock formation geometries.
\Fref{sec:conclusion} outlines conclusions resulting from the numerical study and a discussion of future work.

@@

For the sake of computational simplicity and computational efficiency, the numerical study considered a two-dimensional section of the wellbore and was simulated using the D2Q9 lattice.
In addition, the collision operator used was the BGK collision operator.
For more details on the Lattice Boltzmann Method, the D2Q9 lattice, or the BGK collision operator, see \Fref{sec:LBM}.

@@

Cement slurry is best modeled using a non-Newtonian constitutive relationship.
A simple, yet accurate constitutive relationship for cement slurry is the Bingham plastic constitutive relationship.
Bingham plastics are characterized by a plastic viscosity, which is analogous to the dynamic viscosity for Newtonian fluids, and a yield stress, which is a stress threshold that, if the shear stress does not exceed, the strain-rate is zero.
The Bingham plastic constitutive model is difficult to work with numerically.
Often, a smooth approximation is used instead--the Papastasiou approximation.
Throughout this chapter, the Papastasiou approximation with a stress growth exponent of $m = 10^6$ is used.
The BGK collision operator with $m = 10^6$ can be seen as a compromise between computational efficiency and accuracy for simulating Bingham plastic flows.
For more detail on Bingham plastics or the Papastasiou approximation, see \Fref{sec:bp}.

@@

A free-surface flow is a flow in which two-fluids exist but the dynamics of one fluid does not need to be explicitly modeled.
Typically, the phase of the primary fluid, or the fluid of interest, is a liquid; and the secondary phase that does not need to be explicitly modeled is a gas, often times air at atmospheric pressure.
The primary fluid is modeled using a standard CFD method, such as LBM, and the interaction of the secondary fluid on the primary fluid is modeled as a boundary condition at the interface between the two fluids.
The interface is what is referred to as the free-surface.
The advantage of modeling a multiphase multicomponent flow as a free-surface is that, because the dynamics of the secondary fluid are not explicitly tracked and simulated, the resulting algorithm is more computationally efficient in terms of both memory and CPU overhead.
The algorithm for extending the Lattice Boltzmann Method for free-surface flow used in this work follows the algorithm developed and presented in~\citet{korner2005lattice,thurey2005interactive}.

@@

Consider at each node a cell of size $\text{d}x \times \text{d}x$ where $\text{d}x$ is the distance between nodes and the cell is centered at the node.
In order to track where the primary fluid is and where it is not, i.e. which cells contain the primary fluid, a variable is introduced, $\epsilon$, which describes what fraction of the cell is covered by the primary fluid.
Cells that have a fluid fraction of $\epsilon = 0$ contain no primary fluid, i.e. are empty; cells that have a fluid fraction of $\epsilon = 1$ are filled; and cells that have a fluid fraction $0 < \epsilon < 1$ are neither completely filled nor are they completely empty.
The fluid fraction concept is similar to how fluids are tracked in Volume of Fluid (VOF) methods.

@@

\newcommand{\opidx}{{\bar{i}}}
\newcommand{\fset}{\mathcal{F}}
\newcommand{\iset}{\mathcal{I}}
\newcommand{\gset}{\mathcal{G}}
\newcommand{\domain}{\mathcal{X}}

@@

Each of the cells in the domain are assigned one of three states: fluid, interface, or gas.
The set of fluid, interface, and gas cells are denoted by $\fset$, $\iset$, and $\gset$, respectively.
A fluid cell is a cell that is entirely covered by the primary fluid; and a gas cell is a cell that is completely empty, or does not contain any primary fluid.
An interface is partially covered, and there is always a layer of interface cells between fluid and gas cells so that a fluid cell and a gas cell are never neighbors.
The change in mass is calculated across fluid and interface cells for each discrete velocity direction as follows:
\begin{equation} \label{eq:dM}
\Delta M_i(\pos, t) = \begin{cases}
f_\opidx(\pos + \pvel_i \Delta t, t) - f_i(\pos, t), & \pos \in \fset \text{ or } \pos + \pvel_i \in \fset \\
\left[(\epsilon(\pos, t) + \epsilon(\pos + \pvel_i \Delta t, t)) / 2\right]\left(f_\opidx(\pos + \pvel_i, t) - f_i(\pos, t)\right), & \pos \in \iset \text{ and } \pos + \pvel_i \in \iset \\
0, & \pos + \pvel_i \in \gset
\end{cases}
\end{equation}
\noindent where $\pvel_\opidx$ represents the discrete velocity vector that is opposite in direction to $\pvel_i$, i.e. $\pvel_\opidx = - \pvel_i$.
\eqref{eq:dM} states that when the border of the two cells is covered with fluid, the change in mass for a discrete velocity direction is equal to the difference of the particle distribution entering the cell from the opposite direction and particle distribution leaving the cell in that direction.
When the border of the two cells are not covered with fluid, i.e. when both cells are interface cells, the difference is weighted by the average fluid fraction of the two cells because mass is only able to transfer across the fraction of the cell border that is covered in fluid.
The total change in mass for a cell is simply $\Delta M(\pos, t) = \sum_i \Delta M(\pos, t)$.
Finally, note that $\Delta M_i(\pos, t) = -\Delta M_\opidx(\pos + \pvel_i \Delta t, t)$ for all $\pos \in \domain$ where $\domain$ is the domain.
As a consequence, it is clear that $\sum_{\pos \in \domain}\Delta M(\pos, t) = 0$, i.e. that mass is conserved for the mass transfer step.

@@

\begin{table}
\caption{Cell state transition rules.}
\begin{tabulary}{\linewidth}{l|L|L}
\textbf{Current State} & \textbf{Condition} & \textbf{Transition} \\
\hline &  & \\
$(\pos, t) \in \iset$ & $M(\pos, t) < -\tconst$ & $(\pos, t + \Delta t) \rightarrow \gset$, $M(\pos, t + \Delta t) = 0$ \\
                      & $M(\pos, t) > \tconst + \rho(\pos, t)$ & $(\pos, t + \Delta t) \rightarrow \fset$, $M(\pos, t + \Delta t) = \rho(\pos, t + \Delta t)$ \\
                      &\\
$(\pos, t) \in \fset$ & $(\pos + \pvel_i \Delta t, t + \Delta t) \in \gset$ for any $i = 1, 2, ..., 8$ & $(\pos, t + \Delta t) \rightarrow \iset$ \\
&\\
$(\pos, t) \in \gset$ & $(\pos + \pvel_i \Delta t, t + \Delta t) \in \fset$ for any $i = 1, 2, ..., 8$ & $(\pos, t + \Delta t) \rightarrow \iset$ \\
\end{tabulary}
\label{tab:cell-transition-rules}
\end{table}

\begin{figure}
\centering
\includegraphics{figs/cell-transition}
\caption{Example of cell state transitions.
The interface cell highlighted in red on the left is transitioned to a gas cell as it has emptied.
On the right, the three cells highlighted in red were transitioned from fluid to interface cells in order to keep the interface continuous.}
\label{fig:cell-transition}
\end{figure}

@@

The rules for updating cell states in \Fref{tab:cell-transition-rules} are fairly intuitive, however, there are some bookkeeping nuances that must be considered in order to ensure mass is properly conserved and interface continuity is achieved; e.g. note that if the rules outlined in \Fref{tab:cell-transition-rules} are followed naively that mass is not conserved when interface cells transition to fluid cells or gas cells.
The algorithm for updating cell states follows that which was presented in~\citet{thurey2005interactive}.
It is presented here for the sake of completeness and for outlining minor modifications.
\begin{enumerate}
\item Initialize a variable that represents the global excess mass, e.g. $M_{gex} = 0$.
\item For all cells $\pos \in \iset$, store cells that should be transitioned to $\fset$ in one data structure, $\iset_f$, and store cells that should be transitioned to $\gset$ in another, $\iset_g$.
\item \label{enum:prep-fnbrs} Loop over cells that will be transitioned from interface to fluid cells, i.e. loop over $\iset_f$, and prepare their neighborhoods for transition:
	\begin{enumerate}
	\item Calculate the average density, $\rho_{avg}$, and velocity, $\mathbf{u}_{avg}$, of each neighborhood.
	\item For all cells in the neighborhood of the new fluid cell:
		\begin{enumerate}
			\item If the cell is a gas cell, transition to interface state and set $f_i = f_i^{eq}(\rho_{avg}, \mathbf{u}_{avg})$, $\rho = \rho_{avg}$, and $\mathbf{u} = \mathbf{u}_{avg}$.
			\item If the cell is an interface cell in $\iset_g$, remove it from $\iset_g$.
		\end{enumerate}
	\end{enumerate}
\item \label{enum:prep-gnbrs} Loop over cells that will be transitioned from interface to gas cells, i.e. loop over $\iset_g$, and prepare their neighborhoods for transition:
	\begin{enumerate}
	\item For all cells in the neighborhood of the new gas cell:
		\begin{enumerate}
			\item If the cell is a fluid cell, transition to interface state.
		\end{enumerate}
	\end{enumerate}
\item \label{enum:rloop-fluid} Loop over cells that will be transitioned from interface to fluid cells, i.e. loop over $\pos \in \iset_f$, and:
\begin{enumerate}
	\item Let $M_{ex} = M - \rho$.
	\item Calculate the unit norm outward from the interface, i.e. toward empty space, $\mathbf{n}$.
	Initialize $v_{sum} = 0$.
	\item For all cells in the neighborhood,
	\begin{enumerate}
		\item if an interface cell, store in a data structure, $\iset_{r}$, and for the discrete velocity vector direction, $\pvel_i$, of the respective neighbor, let $v_i = \begin{cases}
					\mathbf{n} \cdot \pvel_i, & \mathbf{n} \cdot \pvel_i > 0 \\
					0, & \text{otherwise}
				\end{cases}$.
		\item $v_{sum} = v_{sum} + v_i$.
	\end{enumerate}
	\item If $v_{sum} \ge 0$, then for each cell, $\pos_j \in \iset_r$, redistribute the excess mass according to $M(\pos_j) = M(\pos_j) + M_{ex} \frac{v_i}{v_{sum}}$. \label{enum:redist-dop-f}
	\item \label{enum:redist-neighbors-f} If $v_{sum} = 0$ and $\iset_r \neq \emptyset$, then for each cell, $\pos_j \in \iset_r$, redistributed the excess mass according to $M(\pos_j) = M(\pos_j) +  M_{ex} / n_r$ where $n_r$ is the number of cells in $\iset_r$.
	\item Otherwise, increment the global excess mass, $M_{gex} = M_{gex} + M_{ex}$.

>>>>

%\usepackage{amsthm}
\usepackage{amssymb}
\usepackage{mathtools}
\usepackage{bm}
\usepackage[plain]{fancyref}
\usepackage{tabulary}
\usepackage{pbox}
\usepackage{algorithm}
\usepackage{algpseudocode}

@@

\tableofcontents
\listoftables
\listoffigures
\listofalgorithms

@@

When oil and gas wellbores are drilled, barriers must be put in place to ensure that fluids do not leak out of the wellbore.
%Wellbore leakage can lead to environmental damage, loss of pressure at the wellhead, and consequently, loss of production.
An important yet vulnerable barrier is the cement annulus and is created through a process known as primary cementing.
Every well has unique subsurface conditions, and so no cement slurry mix design both performs well and is economical for all wells.
%Although some general guidelines and analytical techniques exist for approximating the performance of a cement slurry mix, the mechanics of primary cementing are complex.
Computational methods can help better understand primary cementing and aid designers in determining the optimal mix.
The lattice Boltzmann method (LBM) is a promising technique for simulating primary cementing because it is well-suited for efficiently simulating non-Newtonian flows, multiphase multicomponent flows, and flows through complex geometries--namely, some of the complexities associated with the mechanics of primary cementing.
In \Fref{sec:numerical-methods}, an algorithm for simulating non-Newtonian free-surface flow using LBM is presented.
The algorithm was implemented and used to study primary cementing of a dry annulus, i.e. an annulus that is not filled with drilling mud.
More specifically, the study involved parameterizing different cement slurry flows and investigating how well each slurry flow filled different wellbore geometries.
The study is followed by conclusions and a discussion of future work.

@@

Of critical importance to drilling operations of all kinds is that barriers are placed to prevent gasses and fluids from migrating from one geological zone to another (i.e., zonal isolation is achieved)~\cite{economides19901,jackson1993effect,goodwin1992cement,dusseault2014towards,li2016theory}.
If zonal isolation is not achieved, gasses and fluids may migrate through the well to the surface and up into the atmosphere--potentially causing pollution, or at the very least, contributing to an increase in greenhouse gasses--or, worse yet, formation gasses and fluids can migrate into aquifers, potentially harming wild life and people.
Furthermore, it can be argued that the most important, yet vulnerable, barrier in terms of achieving zonal isolation of a wellbore is the cement annulus that is created between the casing and the formation.
There are various mechanisms by which this cement annulus can fail.
According to previous studies~\cite{Bit02,Che85,Lev79,Ste88,Tal93,Zul12,li2016theory}, the most prevalent failure mechanisms of the cement annulus are:
\begin{itemize}
\item Stresses developing in the cement annulus as a result of temperature gradients, moisture gradients, overburden pressure, and chemical shrinkage of the cement matrix.
These stresses can cause cracking in the cement, or debonding at either the casing--cement or cement--formation interfaces.
\item Poor fill of the annulus.
If cement slurry does not fill the annulus and if drilling mud is left behind the mud can dry, crack, and provide a weak path for fluids and gasses under pressure to push through, or if a dry hole is cemented and not filled properly, fluids and gasses can travel through channels and voids left in the annular space.
\item Gas channeling through the annulus during cement curing.
Initially, after the annulus is poured, the cement column transmits its full hydrostatic pressure against the rock formation.
As the cement annulus hydrates, it solidifies, becoming stiffer and more self-supporting.
When the column begins to support itself, the hydrostatic pressure of the cement column--against the rock formation--begins to drop.
If the hydrostatic pressure of the cement column drops below the pressure of a formation gas before the cement annulus has developed adequate strength to resist it, the formation gas can penetrate into the cement annulus and form channels, or degrade the cement annulus in other ways.
\end{itemize}
% what research has been done that is considered classical?

@@

In order to better understand, and thus improve primary cementing for zonal isolation, several works have attempted to model certain aspects of the multiphase and multiphysics processes of wellbore cement placement.
For example,~\citet{Bei77} developed a mathematical model to describe the displacement of drilling mud by the the cement slurry during primary cementing.
Among their assumptions were that the flow as laminar, the process could be modeled in a quasi-steady manner, the volumetric flow rate was constant at each cross section of the flow, and the pressure fields were only dependent on the distance from the center of the annulus to either the rock formation or steel casing surface.
The model was then used to explore the effect of various parameters on the displacement efficiency of the spacer, such as the density ratio between the mud and the spacer, the viscosity ratio between the mud and the spacer, the displacement rate, the displaced phase yield stress, and the displacing phase yield stress.
It was concluded that displacement efficiency could be improved by increasing the density of the displacing phase, increasing the viscosity of the displacing phase, (in general) decreasing the displacement rate, decreasing the yield stress of the displaced phase, and increasing the yield stress of the displacing phase.
\citet{Bit02} developed a mathematical model for primary cementing of an oil well using a Hele-Shaw displacement model.
The model was more general than the model developed by~\citet{Bei77} as it used less restrictive assumptions and was able to consider an eccentric annulus.
The bulk fluid motion of the spacer and drilling mud during primary cementing were investigated.
The intended use of the model was not to make general statements about what properties of the spacer and drilling mud were desirable, but to instead be used iteratively during the design process for the cement job.
Results showed that the model was able to simulate an unyielded channel of mud left on the narrow side of the annulus.
\citet{Zul12} used a computational fluids dynamics (CFD) solver with volume of fluid (VOF) method to simulate the cementing process of a 50 ft. section of wellbore.
The model included the interfacial dynamics of, not just the spacer and drilling mud, but also the cement slurry--which was not considered in either~\citet{Bei77} or~\citet{Bit02}.
The drilling mud and cement slurry properties were kept constant, while the spacer density, viscosity, and displacement rates were varied.
\citet{Zul12} concluded that the high displacement efficiency occurred when the spacer was the same density as the drilling mud and had a smaller viscosity, similar to that of water.
The displacement efficiency was poor for all scenarios in which the spacer and cement slurry densities were equal.
\citet{zhao2016lattice} developed a model simulating primary cementing using the Lattice Boltzmann Method.
\citet{zhao2016lattice} used the model to investigate the shape of the interface between the displacing phase (i.e. cement slurry) and displaced phase (i.e. drilling mud) in a horizontal, eccentric annulus.
It was found, in general, that a greater density ratio between the displacing phase and displaced phase and lower eccentricity results in better displacement in a horizontal, eccentric annulus.

@@

The purpose of the current work is to develop a computational approach to investigate the flow of cement slurry in and around the--possibly--imperfect features of a rock formation in a vertical section of a wellbore, with the hope of better understanding what parameters of the primary cement job will lead to a good fill of the cement annulus and high bond strength at imperfect rock formation features.
The Lattice Boltzmann Method (LBM) was chosen as the most suitable modeling technique for the desired purpose because of its ability to model flow through and around complex geometries~\cite{thorne2006lattice}, such as those found at the surface of the rock formation.
LBM also has other advantages as an approach for simulating primary cementing:
\begin{itemize}
\item LBM is well-suited for simulating non-Newtonian fluids because the strain-rate tensor is computed local to each node and is second-order accurate~\cite{kruger2009shear,kruger2010second}.
Many fluids present in primary cementing such as the spacer, drilling mud, and most importantly, the cement slurry, are best modeled as non-Newtonian fluids.
\item LBM is well-suited for simulating multiphase multicomponent flows and free-surface flows because it can handle complicated interface shapes between fluid phases and components.
\item LBM is easily written in parallel.
As hardware architectures shift from one or a few cores to several CPUs, it becomes increasingly important that code for simulating complex and computationally expensive phenomena be written in parallel.
\end{itemize}
The current work focuses on the problem of cementing a dry hole, i.e. an annulus that is not filled with drilling mud prior to cementing.
In order to numerically investigate this problem, an LBM model for non-Newtonian and free-surface flow is developed.
The details of the LBM model are outlined in \Fref{sec:numerical-methods}.
What follows in \Fref{sec:results} is a preliminary study on how different characteristics of flow, e.g. cement slurry yield stress, cement slurry inlet velocity, etc., perform in regards to fill of the annulus for different wellbores with imperfect rock formation geometries.
\Fref{sec:conclusion} outlines conclusions resulting from the numerical study and a discussion of future work.

@@

For the sake of computational efficiency, the numerical study considered a two-dimensional section of the wellbore and was simulated using the D2Q9 lattice.
Idealizing the annulus as two-dimensional limits the modeling approach in that annulus eccentricity and other three-dimensional phenomena could not be considered.
A two-dimensional representation of the annulus was deemed sufficient as the focus of this study is to establish the Lattice Boltzmann Method as a suitable method for simulating primary cementing in an annulus with an imperfect rock formation surface.
In addition to using a D2Q9 lattice, the BGK collision operator used was in the current chapter.
For more details on the fundamentals of the Lattice Boltzmann Method, the D2Q9 lattice, or the BGK collision operator, see \Fref{sec:LBM}.

@@

Cement slurry is typically modeled using a non-Newtonian constitutive relationship.
%A simple, yet accurate constitutive relationship for cement slurry is the Bingham plastic constitutive relationship.
As is commonly utilized for modeling cement and concrete fluid dynamics~\cite{ferraris2001fresh,ferraris1998testing,tattersall2003workability,roussel2007computational,roussel2015numerical,svec2012free}, the Bingham plastic constitutive relationship was utilized herein.
Bingham plastics are characterized by a plastic viscosity, which is analogous to the dynamic viscosity for Newtonian fluids, and a yield stress, which is a stress threshold that, if the shear stress does not exceed, the strain-rate is zero.
The Bingham plastic constitutive model is difficult to work with numerically.
Therefore, a smooth approximation, specifically the Papastasiou approximation, is often used~\cite{papanastasiou1987flows,chen2014simulations,tang2011bingham,Cha11,wang2008lattice,phillips2011lattice} and is used herein.
Throughout this chapter, the Papastasiou approximation with a stress growth exponent of $m = 10^6$ is used.
The BGK collision operator with $m = 10^6$ can be seen as a compromise between computational efficiency and accuracy for simulating Bingham plastic flows.
For more detail on Bingham plastics or the Papastasiou approximation, see \Fref{sec:bp}.

@@

A free-surface flow is a flow in which two-fluids exist but the dynamics of one fluid does not need to be explicitly modeled.
Typically, the phase of the primary fluid, or the fluid of interest, is a liquid; and the secondary phase that does not need to be explicitly modeled is a gas, often times air at atmospheric pressure.
The primary fluid is modeled using a standard CFD method, such as LBM, and the interaction of the secondary fluid on the primary fluid is modeled as a boundary condition at the interface between the two fluids.
The interface is what is referred to as the free-surface.
The advantage of modeling a multiphase multicomponent flow as a free-surface is that, because the dynamics of the secondary fluid are not explicitly tracked and simulated, the resulting algorithm is more computationally efficient in terms of both memory and CPU overhead.
The algorithm for extending the Lattice Boltzmann Method for free-surface flow used in this work follows the algorithm developed and presented in~\citet{korner2005lattice} and \citet{thurey2005interactive}.

@@

Consider at each node a cell of size $\text{d}x \times \text{d}x$ where $\text{d}x$ is the distance between nodes and the cell is centered at the node.
In order to track where the primary fluid is and where it is not (i.e. which cells contain the primary fluid), a variable is introduced, $\epsilon$, which describes what fraction of the cell is covered by the primary fluid.
Cells that have a fluid fraction of $\epsilon = 0$ contain no primary fluid (i.e. are empty); cells that have a fluid fraction of $\epsilon = 1$ are filled; and cells that have a fluid fraction $0 < \epsilon < 1$ are neither completely filled nor are they completely empty.
The fluid fraction concept is similar to how fluids are tracked in Volume of Fluid (VOF) methods.

@@

\newcommand{\opidx}{{\bar{\i}}}
\newcommand{\fset}{\mathcal{F}}
\newcommand{\iset}{\mathcal{I}}
\newcommand{\gset}{\mathcal{G}}
\newcommand{\domain}{\mathcal{X}}

@@

Each of the cells in the domain are assigned one of three states: fluid, interface, or gas.
The set of fluid, interface, and gas cells are denoted by $\fset$, $\iset$, and $\gset$, respectively.
A fluid cell is a cell that is entirely covered by the primary fluid; and a gas cell is a cell that is completely empty, or does not contain any primary fluid.
An interface is partially covered, and there is always a layer of interface cells between fluid and gas cells so that a fluid cell and a gas cell are never neighbors.
The change in mass is calculated for fluid and interface cells (i.e. $\pos \in \fset \cup \iset$) for each discrete velocity direction as follows:
\begin{equation} \label{eq:dM}
\Delta M_i(\pos, t) = \begin{cases}
f_\opidx(\pos + \pvel_i \Delta t, t) - f_i(\pos, t), & \pos \in \fset \text{ or } \pos + \pvel_i \Delta t \in \fset \\
%\left[(\epsilon(\pos, t) + \epsilon(\pos + \pvel_i \Delta t, t)) / 2\right]\left(f_\opidx(\pos + \pvel_i, t) - f_i(\pos, t)\right), & \pos \in \iset \text{ and } \pos + \pvel_i \Delta t \in \iset \\
\frac{\epsilon(\pos, t) + \epsilon(\pos + \pvel_i \Delta t, t}{2} (f_\opidx(\pos + \pvel_i, t) - f_i(\pos, t)), & \pos \in \iset \text{ and } \pos + \pvel_i \Delta t \in \iset \\
0, & \pos + \pvel_i \Delta t \in \gset
\end{cases}
\end{equation}
\noindent where $\pvel_\opidx$ represents the discrete velocity vector that is opposite in direction to $\pvel_i$ (i.e. $\pvel_\opidx = - \pvel_i$).
\eqref{eq:dM} states that when the border of the two cells is covered with fluid, the change in mass for a discrete velocity direction is equal to the difference of the particle distribution entering the cell from the opposite direction and particle distribution leaving the cell in that direction.
When the border of the two cells are not covered with fluid (i.e. when both cells are interface cells) the difference is weighted by the average fluid fraction of the two cells because mass is only able to transfer across the fraction of the cell border that is covered in fluid.
The total change in mass for a cell is simply $\Delta M(\pos, t) = \sum_i \Delta M_i(\pos, t)$.
Finally, note that $\Delta M_i(\pos, t) = -\Delta M_\opidx(\pos + \pvel_i \Delta t, t)$ for all $\pos \in \domain$ where $\domain$ is the domain.
As a consequence, it is clear that $\sum_{\pos \in \domain}\Delta M(\pos, t) = 0$ (i.e. that mass is conserved for the mass transfer step).

@@

\begin{table}
\caption{Cell state transition rules.}
\begin{tabulary}{\linewidth}{l|L|L}
\textbf{Current State} & \textbf{Condition} & \textbf{Transition} \\
\hline & \\
$(\pos, t) \in \iset$ & $M(\pos, t) < -\tconst$ & $(\pos, t + \Delta t) \rightarrow \gset$, $M(\pos, t + \Delta t) = 0$ \\
                      & $M(\pos, t) > \tconst + \rho(\pos, t)$ & $(\pos, t + \Delta t) \rightarrow \fset$, $M(\pos, t + \Delta t) = \rho(\pos, t + \Delta t)$ \\
                      & &\\
$(\pos, t) \in \fset$ & $(\pos + \pvel_i \Delta t, t + \Delta t) \in \gset$ for any $i = 1, 2, ..., 8$ & $(\pos, t + \Delta t) \rightarrow \iset$ \\
& & \\
$(\pos, t) \in \gset$ & $(\pos + \pvel_i \Delta t, t + \Delta t) \in \fset$ for any $i = 1, 2, ..., 8$ & $(\pos, t + \Delta t) \rightarrow \iset$ \\
\end{tabulary}
\label{tab:cell-transition-rules}
\end{table}

\begin{figure}
\centering
\includegraphics[width=\linewidth]{figs/cell-transition}
\caption{Example of cell state transitions.
The interface cell highlighted in red on the left is transitioned to a gas cell as it has emptied.
On the right, the three cells highlighted in red were transitioned from fluid to interface cells in order to keep the interface continuous.}
\label{fig:cell-transition}
\end{figure}

@@

The rules for updating cell states in \Fref{tab:cell-transition-rules} are fairly intuitive, however, there are some details that must be considered in order to ensure mass is properly conserved and interface continuity is achieved (e.g. note that if the rules outlined in \Fref{tab:cell-transition-rules} are followed naively that mass is not conserved when interface cells transition to fluid cells or gas cells).
The algorithm for updating cell states follows that which was presented in~\citet{thurey2005interactive}.
It is presented here as \Fref{algo:update-cell-states} for the sake of completeness and for outlining minor modifications.

\input{src/algorithm/update-cell-states.tex}

\Fref{algo:update-cell-states}, as presented, ensures both mass conservation and continuity of the free-surface.
Moreover, when mass redistribution inevitably occurs, it is first attempted to be redistributed and weighted in the direction in which the interface is propagating (e.g. line \algref{algo:update-cell-states}{line:redist-dop-f} or line \algref{line:redist-dop-g}.
If there are no interface cells in the direction of interface propagation, it is instead redistributed to neighboring interface cells (e.g. line \algref{line:redist-neighbors-f} or line \algref{line:redist-neighbors-g}).
If there is no neighboring cells which are interface cells, the excess mass is added onto the global excess mass, $M_{gex}$, which is eventually redistributed among all of the interface cells (e.g. line \algref{line:final-redist}).
The hierarchy of possible redistribution steps ensures that mass redistribution is done in the most physically meaningful way as is possible. % We may need to hard code references to steps in here...
Lastly, note that the interface normal, $\mathbf{n}$, is calculated using a finite difference approximation,
\begin{equation} \label{eq:n}
\mathbf{n}_{ij} = \frac{1}{2} \begin{Bmatrix}
\epsilon_{i-1, j} - \epsilon_{i+1, j} \\
\epsilon_{i, j-1} - \epsilon_{i, j+1}
\end{Bmatrix}
\end{equation}
\noindent where $i$ is the index of the node in the x-direction and $j$ is the index of the node in the y-direction.
